---
title: "TIME SERIES"
author: "Hugo S. ROCHA MONDRAGON"
date: "2025-03-07"
output: 
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PART 1 : Nelson-Siegel Model

### Explanation and plotting
The Nelson-Siegel model is a parametric model for fitting and forecasting the yield curve of interest rates. It provides a representation of the term structure of interest rates using an exponential decay function.

$$
y(\tau) = \beta_0 + \beta_1 \frac{1 - e^{-\tau/\lambda}}{\tau/\lambda} + \beta_2 \left( \frac{1 - e^{-\tau/\lambda}}{\tau/\lambda} - e^{-\tau/\lambda} \right)
$$
Where: 

-   $y(\tau)$ represents the yield at maturity $\tau$
-   $\beta_0$, the intercept, represents the long-term level of the yield curve, it captures the general level of interest rates across all maturities
-   $\beta_1$, the slope coefficient, determines the slope of the yield curve, it mainly affects the difference between short-term and long-term interest rates
-   $\beta_2$ controls the curvature of the yield curve, it is most influential in the medium-term maturities
-   $\lambda$ determines the speed at which the impact of $\beta_1$ and $\beta_2$ fades as maturity $\tau$ increases

We have already implemented a function (NS) based on the Nelson-Siegel model. Let's begin by calling it.
```{r}
source("C:\\Users\\Sebastian\\OneDrive - Université Paris-Dauphine\\DAUPHINE\\01_M1\\S2\\Info\\NS_fct.R")
```

To store our time-to-maturity values, we define a vector t, representing $\tau$ The numbers 1, 3, 6, etc., correspond to maturities in months, which we convert into years by dividing by 12.
```{r}
t<-( c(1,3,6,9,12,15,18,21,24,30,36,48,60,72,84,96,108,120)/12)
```

We execute our function, ensuring that the chosen parameters produce a normal yield curve. To achieve this, I set $\beta_1$ (short-term effect) smaller than $\beta_2$ (medium-term hump effect), which in turn is smaller than $\beta_0$ (long-term level).
```{r}
NS_1 <- NS(2.05,-1.80,2.1,0.85,t)
```

```{r}
layout(matrix(1:1,1,1))
plot.ts(NS_1,ylab="zc rate", main="Yield curve")
```


### Extension on $\beta_1$ and $\beta_2$ parameters

As previously stated:

-  $\beta_1$ determines the **slope** of the yield curve, primarily affecting short-term rates.
-  $\beta_2$ controls the **curvature**, influencing mid-term rates by introducing a hump effect.

We will now analyze how the yield curve reacts to changes in these parameters.

#### Effect of $\beta_1$ (Slope Parameter)

We create a loop to observe how modifying $\beta_1$ impacts the yield curve.

```{r}
layout(matrix(1:6, ncol = 2))
B0 <- 2.05
B2 <- 2.1
L  <- 0.85
B1 <- -1.8
for (i in 1:6) {
  B1 <- B1 + 0.25  
  yield_curve <- NS(B0, B1, B2, L, t) 
  plot(t, yield_curve, type = "l", col = "blue", lwd = 2, 
       xlab = "Maturity (Years)", ylab = "Yield (%)",
       main = paste("Yield Curve for B1 =", round(B1, 2)))
  grid()
}
```

**Observations on $\beta_1$:**

As we increase $\beta_1$: 

-  The slope of the yield curve increases, meaning short-term rates become lower relative to long-term rates.
-  A higher $\beta_1$ leads to a steeper yield curve, consistent with a normal yield curve.

#### Effect of $\beta_2$ (Curavature Parameter)

```{r}
layout(matrix(1:6, ncol = 2))
B0 <- 2.05
B1 <- -1.8  
L  <- 0.85
B2 <- 2.1
for (i in 1:6) {
  B2 <- B2 + 0.5  # Increment B2
  yield_curve <- NS(B0, B1, B2, L, t)  # Compute yield curve
  plot(t, yield_curve, type = "l", col = "red", lwd = 2, 
       xlab = "Maturity (Years)", ylab = "Yield (%)",
       main = paste("Yield Curve for B2 =", round(B2, 2)))
  grid()
}
```

**Observations on $\beta_2$:**

As we increase $\beta_2$: 

-  The hump in the mid-maturity range(2-5 years) becomes more pronounced
-  A higher $\beta_2$ introduces more curvature, making medium-term rates deviate more from short and long-term rates.

## PART 2 : Nelson-Siegel-Svensson Model

### Explanation and Plotting

The **Nelson-Siegel-Svensson model** is an extension of the Nelson-Siegel model, adding an extra term to better fit the yield curve, particularly allowing for a second hump.

$$
y(\tau) = \beta_0 + \beta_1 \frac{1 - e^{-\tau/\lambda_1}}{\tau/\lambda_1} + \beta_2 \left( \frac{1 - e^{-\tau/\lambda_1}}{\tau/\lambda_1} - e^{-\tau/\lambda_1} \right) + \beta_3 \left( \frac{1 - e^{-\tau/\lambda_2}}{\tau/\lambda_2} - e^{-\tau/\lambda_2} \right)
$$

Where:  

-  $y(\tau)$ represents the yield at maturity $\tau$.
-  $\beta_0$, the intercept, represents the **long-term level** of the yield curve.
-  $\beta_1$, the slope coefficient, determines the **steepness** of the curve.
-  $\beta_2$, the curvature parameter, influences the **medium-term hump**.
-  $\beta_3$, the second curvature parameter, adds **additional flexibility**, allowing a second hump or dip.
-  $\lambda_1$ and $\lambda_2$ determine how quickly the effects of $\beta_1$, $\beta_2$, and $\beta_3$ decay as maturity increases.



Note that our NSS function is contained in the same file as the NS function so we dont need to call it again. 


```{r}
NSS_1 <- NSS(2.05,-1.80,2.1,0.5,0.85,1.5,t)
layout(matrix(1:1,1,1))
plot.ts(NSS_1, ylab="zc rate", main="Yield Curve (NSS Model)")
```


### Extension on $\beta_3$ parameter

As previously stated:

-  $\beta_1$ determines the **slope** of the yield curve, primarily affecting short-term rates.
-  $\beta_2$ controls the **curvature**, influencing mid-term rates by introducing a hump effect.
-  $\beta_3$ allows a **second curvature**, allowing for more flexibility.

As we already analyzed $\beta_1$ and $\beta_2$ in part 1, we will only analyze $\beta_3$ now. 

#### Effect of $\beta_3$ (Second Curvature Parameter)


```{r}
layout(matrix(1:6, ncol = 2))
B0 <- 2.05
B1 <- -1.8  
B2 <- 2.1
L1  <- 0.85
L2  <- 1.5
B3 <- 0.5

for (i in 1:6) {
  B3 <- B3 + 0.5  
  yield_curve <- NSS(B0, B1, B2, B3, L1, L2, t)  
  plot(t, yield_curve, type = "l", col = "purple", lwd = 2, 
       xlab = "Maturity (Years)", ylab = "Yield (%)",
       main = paste("Yield Curve for B3 =", round(B3, 2)))
  grid()
}
```

**Observations on $\beta_3$:**

-  Increasing $\beta_3$ adds a scond hump in long maturities.
-  Higher $\beta_3$ increases yield curve flexibility.
